Analysing the Performance of Shallow-Water Models with the Roofline Model

### 2016-02-23 docs/papers/roofline_model/gocean_roofline.tex
roofline: text describing SIMD opts and update of AVX performance figures

>>>>

Examining the optimisation report produced by the Intel compiler
revealed that the majority of array accesses within the loop were not
aligned:
\begin{verbatim}
Example of unaligned access report here
\end{verbatim}
and that this inhibited efficient vectorisation. Tackling this issue
requires two steps; first, ensure that the arrays are allocated to
align with 64-byte boundaries and second, ensure that the compiler is
aware of this where those arrays are used. The first requirement is
easily met by passing the ``-align array64byte'' flag to
the compiler when building the code. This ensures that any array that
is constructed using a call to ALLOCATE is aligned on 64-byte
boundaries in memory.

The second requirement is complicated by the fact that the NEMOLite2D
code makes extensive use of Fortran derived types --- the data arrays
used for computation are members of derived types. The information
about the alignment of these arrays cannot therefore be propagated
through the code by module {\tt use} statements. In fact, even the
Intel compiler directive which can be used to specify the alignment
of a particular array at its point of use will not accept references
to members of a derived type. It was therefore necessary to wrap the
kernel call (and associated loops) within a routine and pass in
references to the arrays. Within the wrapper routine the corresponding
dummy arguments are declared as simple arrays and their alignment can
then be specified using the {\tt ASSUME\_ALIGNED} directive:
\begin{verbatim}
subroutine invoke_continuity_arrays(nx, ny, ssha, ...)
  integer, intent(in)   :: nx, ny
  real(wp), intent(out) :: ssha(nx,ny)
!DIR$ ASSUME_ALIGNED ssha:64
  ...
\end{verbatim}
This approach also enables the extent of the arrays ({\tt nx} and {\tt
  ny} in this example) to be specified explicitly.

The continuity kernel includes array accesses of the form {\tt
  a(i,j-1)}. If array {\tt a} is allocated with extent {\tt (Nx, Ny)}
then this specifies a location in memory {\tt Nx} words before that
occupied by {\tt a(i,j)}. In order for such an access to be aligned,
{\tt Nx} must be a multiple of the SIMD vector width (2 for SSE, 4 for
AVX) and this must be specified to the compiler. We therefore padded Nx
as required and added a directive within the wrapper routine:
\begin{verbatim}
!DIR$ ASSUME (MOD(Nx,ALIGNMENT) .EQ. 0)
\end{verbatim}

Once all of these steps were complete, the only un-aligned array
accesses reported by the compiler were those of the form {\tt
  a(i-1,j)}. Without taking a physical copy of the array {\tt a}, such
unaligned accesses are unavoidable in a stencil-based
finite-difference code.

The final SIMD-related optimisations are concerned with what happens
at the beginning and end of a SIMD-vectorisable loop. If the loop
updates e.g. {\tt a(i,j)} and begins on a value of {\tt i} that
  results in unaligned memory accesses (e.g. {\tt i = 2}) then the
  compiler must insert a ``peel'' loop to handle those iterations
  required to reach a point where memory is accessed on a 64-byte
  boundary.  Similarly, if the trip-count of the resulting
  vectorisable loop is not an integer multiple of the SIMD vector
  width then the compiler must insert a ``remainder'' loop to handle
  those iterations left over.  We removed the remainder loop by
  ensuring that the length of the inner loop was a multiple of the
  vector width and adding an {\tt ASSUME} compiler directive
  (c.f. that for {\tt Nx} above). This was straightforward since we
  had already padded the array extents themselves and NEMOLite2D uses
  a mask to define the active computational domain - the additional
  grid points resulting from the padding were simply marked as
  external to the domain.  Despite doing this we found that the
  remainder loop was not removed until we manually introduced the peel
  loop (rather than relying on the compiler).

The inner loop of the kernel where all array accesses are aligned and
the number of distinct streams has been reduced to eight (with a
concommitant reduction in the number of FLOPs):
\begin{verbatim}
do ji = ALIGNMENT+1, M, 1
   rtmp1 = (sshn_u(ji  ,jj ) + hu(ji  ,jj  ))*un(ji  ,jj)
   rtmp2 = (sshn_u(ji-ALIGNMENT,jj ) + hu(ji-ALIGNMENT,jj))* &
           un(ji-ALIGNMENT,jj)
   rtmp3 = (sshn_v(ji ,jj )  + hv(ji  ,jj  ))*vn(ji ,jj)
   ssha(ji,jj) =  (rtmp2 - rtmp1 - rtmp3) / sshn_t(ji,jj)
end do
\end{verbatim}
